Kernel Phase and Coronagraphy with Automatic Differentiation

### 2020-07-27 ms.tex

%% Mark off the abstract in the ``abstract'' environment. 
\begin{abstract}
% optical imaging with aberrations
Aberrations in the optical path introduce distortions and speckles in the resulting images which limit the performance of cameras at high angular resolution.

% self cal - closure phase and kernel phase
Self-calibrating observables such as the `closure phase' or `bispectrum' have been widely used in optical and radio astronomy to correct for this: a linear combination of measurements with phase noise can be made such that error terms cancel, leaving a subspace of high signal-to-noise observables. Kernel phases are a generalization of closure phases, in which the effect of aberration on an arbitrary pupil is approximated as a linear phase transfer matrix.

%coronagraphs are hard
Finding this operator is analytically tractable between a single pupil and focus but is much more difficult for other optical systems of interest, such as coronagraphs.
% autodiff can make this practical for arbitrary optical systems
Automatic differentiation or `backpropagation' software now makes calculating derivatives with respect to aberrations in arbitrary planes straightforward for any optical system.

% we demonstrate a simple example of the kernel of a monochromatic coronagraph
We reproduce existing kernel phase theory within this framework and demonstrate an extension to the case of a Lyot coronagraph, which is found to have self-calibrating combinations of speckles which are resistant to phase noise, but only in the very high wavefront quality regime. We also use this framework to reanalyze Palomar adaptive optics observations of the binary $\alpha$~Ophiuchi, finding broad consistency and slight systematics between the new pipeline and the existing standard.

%phase design
We present a Python package \textsc{morphine}, with an interface similar to the popular package \textsc{poppy}, for optical simulation with automatic differentiation. These methods may be useful for designing improved astronomical optical systems by gradient descent.
\href{https://github.com/benjaminpope/morphine}{\color{linkcolor}\faGithub} % shamelessly borrowed from Luger!

@@

\section{Introduction} 
\label{sec:intro}
% optical imaging with aberrations
Many questions in astronomy and other sciences can only be answered with diffraction-limited high resolution imaging. The highest resolutions are typically achieved with the method of interferometry, in which waves detected at multiple receivers are combined physically or in post-processing to obtain the Fourier transform of the source intensity distribution \citep{vc34,zernike38}. Even in the case of a single telescope or camera, it is often nevertheless helpful to think of them as an interferometer composed of many sub-apertures which combine their signals directly onto a focal plane \citep[a `Fizeau interferometer':][]{fizeau1868}}. In all of these cases, unknown path delays toward each receiver or subaperture are a dominant source of noise, causing `tip-tilts' of the final image position, distortions of the point spread function (PSF) from low-spatial-frequency aberrations, and clouds of `speckles' from high-spatial-frequency aberrations.

% self cal with closure phase
The traditional answer to this has been self-calibration with `closure phases' \citep[introduced in the context of radio astronomy by][]{jennison58}}, in which phases are added around three interferometric baselines which form a closing triangle. The phase error terms local to each subaperture cancel, so that you trade three low-signal-to-noise (SNR) baseline phases for one high-SNR observable. In a non-redundant array (one in which no baseline is repeated between different pairs of subapertures) of sufficient size, a large number of closure phases can be obtained which anchor image reconstruction with great precision. `Closure amplitudes' which are resistant to fluctuations in input amplitude or gain can also be obtained by a similar construction using four telescopes \citep{twiss60}. The closure phase is the argument of a quantity called the triple product or `bispectrum', and it has recently been shown from the perspective of invariant theory in algebraic geometry that for a wide class of problems limited by phase noise, knowledge of the mean, power spectrum and bispectrum are necessary and sufficient for an optimal signal reconstruction \citep{bandeira17}.

On the other hand, in exoplanet direct imaging with coronagraphs, analytic self calibrations such as this are not yet known. The standard approaches to data analysis in coronagraphy rely on exploring a diversity of PSFs experimentally, constructing a basis covering some of the diversity in speckle patterns, and then subtracting out a linear combination of vectors in this basis from measured data \citep[e.g.][]{lafreniere07,soummer12,pueyo16}}. Advantages can also be gained from angular differential imaging \citep[ADI;][]{marois06}} and spectral deconvolution with wavelength \citep{sparks02}. In this paper we will extend our understanding of analytic self calibration to better include coronagraphs, so that we can add an additional layer of precision to exoplanet imaging calibration.

@@

% coronagraphs, speckles etc
It is desirable to find a priori self-calibration schemes for high-angular-resolution, high-contrast imaging systems more generally, including not merely nulling interferometers but optically complex coronagraphs, which block out light in a region around a target star. One particularly interesting extension of the kernel phase idea is the `kernel nuller' concept \citep{martinache18}, in which the aberration transfer matrix and kernel operator idea is applied to the output of a nulling interferometer.

@@

The \textsc{DeepOptics} project \citep{sitzmann2018} has used autodiff to optimize `computational cameras'. Building their model in \textsc{TensorFlow}, they couple a physical optics simulation, detector simulation, and a deconvolution post-processing stage for a total end-to-end imaging simulation. Where normally you might optimize some heuristic of the PSF (such as the full width at half maximum), this makes it possible to jointly optimize hardware and software with respect to figures of merit of the overall system such as final resolution or depth of field. Some designs this way are exotic: for example, super-resolution is achieved by finding a lens with three off-centre Fresnel-lens components focusing to three separate spots, which the deconvolution stage shifts and stacks. Because it is built in \textsc{TensorFlow}, the diffraction simulation can be incorporated as a `physical layer' in neural network applications in microscopy, for example for optimizing hardware and software for image classification \citep{muthumbi19}, or with reinforcement learning for adaptively learning sample illumination \citet{chaware19}.

@@

Rather than using these autodiff gradients for optimization, in this paper we want to use them to understand optical systems and the information they propagate.
In the following we will show how autodiff can reproduce the existing state of the art in kernel phase, and demonstrate a way forward using autodiff to extend the kernel phase idea to coronagraphic instruments.

@@

For our forwards model, we adapt the popular physical optics library \textsc{poppy} \citep{poppy}, with the goal of compatibility with existing simulations built on \textsc{poppy} such as \textsc{WebbPSF} \citep{webbpsf}. We use the matrix Fourier transform mode to avoid using the FFT \citet{soummer07}, which is more efficient for autodiff applications. We have adapted the low-level features of \textsc{poppy} to use the Google autodiff library \textsc{jax} \citep{jax} in place of \textsc{NumPy}, and to distinguish it from the original version we call this new \textsc{poppy} derivatives library \textsc{morphine}. %The \textsc{NumPy}-like API of \textsc{jax} is chosen so that astronomers can painlessly integrate \textsc{morphine} with their existing code, but we note that a \textsc{DeepOptics}-like approach using \textsc{TensorFlow} may be preferable in the longer term for integration with neural networks.

@@

First we want to reproduce the basic kernel phase calculations in this new model. We propagate monochromatic 2.0\,$\mu$m light through a 2.0\,m diameter circular pupil onto a detector with a 20\,mas/pixel scale and 4\,arcsec field of view. The pupil and $u,v$ are each calculated on a 64x64 grid. We evaluate the Jacobian of the $u,v$ phases with respect to the input pupil phases using forwards-mode autodiff and then evaluate this Jacobian at zero phase to get our $\mathbf{A}$ matrix\footnote{\href{https://github.com/benjaminpope/morphine/blob/master/notebooks/morphine_u,v.ipynb}{github.com/benjaminpope/morphine/blob/master/notebooks/morphine\_u,v.ipynb}}. On a laptop computer this takes $\sim$ a few minutes. Memory usage is the main overhead here: in practice on a laptop with 16GB RAM, we are limited to grids no larger than 128x128. This is probably the most major limitation of the method; in future it may be overcome by more efficient refactoring of the code, or simply by doing calculations on clusters with very large RAM allocations.

It is inefficient to calculate this full Jacobian with respect to a densely-pixelized basis for the wavefront in the pupil plane, but we do so here for completeness: in practical cases, aberrations are typically dominated by low-order modes and we could consider the Jacobian only with respect to (for example) some much smaller number of Zernike modes. Reducing the input dimensionality like this would have significant advantages in memory using for forwards-mode autodiff.

@@

We then extract kernel phases from both $\alpha$~Oph and a point source calibrator $\eps$~Her using a version of \textsc{xara} modified to use the \textsc{morphine} matrix FT in place of its own. Following \citet{martinache20}, we choose as our observables the median kernel phases across all 100 exposures for each, and the standard error of the mean as our base uncertainty on each. To calibrate the kernel phases, we simply subtract those of the calibrator from $\alpha$~Oph and add their uncertainties in quadrature.

To see the effect of the `ghost' introduced by the neutral density filter used to observe such bright stars, we calculate kernel phase `colinearity' maps as in \citet{martinache20}, shown in Figure~\ref{colinearity}. In these maps we can see that the filter ghost shows up strongly as a false binary in the uncalibrated kernel phase maps, but calibrated kernel phases easily remove this and reveal the tight $\alpha$~Oph binary.

@@

We then use Markov Chain Monte Carlo \citep{metropolis53} to sample from the posterior assuming Gaussian likelihoods, using the \texttt{emcee} affine-invariant ensemble sampler \citep{emcee}. We obtain binary parameters of separation $\rho = 115.3 \pm 1.2$~mas, position angle $\theta = 85.6 \pm 0.27$~degrees, and contrast $c = 26.3 \pm 0.6$: these are close to the \citet{martinache20} figures, but differ by much more than their small uncertainties. For comparison, the contrast ratio of the system is expected from resolved imaging to be $28.0 \pm 8.3$ in $K$-band \citep{hinkley11}. Posteriors for both inferences are shown in Figure~\ref{comparison_posterior}.

@@

In order to investigate this discrepancy, we generated a grid of forwards models. Using the same code used to produce the kernel phase autodiff model, we generated input wavefronts with 20 Zernike modes of aberrations, each with random amplitudes normally-distributed with 15\,nm standard deviation, for a total peak-to-peak amplitude of 206\,nm. These were then shifted and added to simulate a range of binaries at different configurations and contrasts, and both the \citet{martinache20} pipeline and the autodiff pipeline used to retrieve their parameters. The results are shown in Figure~\ref{method_comparison_ensemble}. The pipelines achieve broadly comparable performance on these simulations, with the \citet{martinache20} method being slightly better at closer separations and higher contrasts and autodiff performing marginally better at larger separations.

@@

\begin{figure}
\plotone{comparison_posterior_sim.pdf}
\caption{Comparison of \citet[][blue stars]{martinache20} and autodiff (orange dots) kernel phase pipelines on simulated data with true parameters $\rho = 125$~mas, $\theta = 86^\circ$, $c=25$. We generate a cube of 1000 mock datasets using a 64-pix pupil model and Zernike aberrations as described in Section~\ref{sec:palomar}, and the observables used are the median and standard error of the mean of each kernel phase across this ensemble. Both datasets retrieve the true parameters very precisely, though with small systematic offsets, which are slightly larger for the autodiff method than for classic kernel phases. \label{comparison_posterior_sim}}
\end{figure}

\begin{figure}
\plotone{comparison_posterior_sim_cube.pdf}
\caption{Comparison of \citet[][blue stars]{martinache20} and autodiff (orange dots) kernel phase pipelines on simulated data with true parameters $\rho = 125$~mas, $\theta = 86^\circ$, $c=25$. We generate a cube of 1000 mock datasets using a 64-pix pupil model and Zernike aberrations as described in Section~\ref{sec:palomar}. The corner plot displays the parameters retrieved using a simple least squares fit applied to each individual frame. The methods have very similar distributions centred tightly on the true value. \label{comparison_posterior_sim_cube}}
\end{figure}

In numerical experiments, subpixel mismatch in Fourier transform coordinates or even Fourier centering convention could cause mismatches between simulation parameters and retrievals at around the scale of the discrepancy between the two methods.  These results are in line with the \citet{martinache_habilitation} conclusion that inexact Fourier sampling contributes significant systematic noise to kernel phases.The final version used for the autodiff pipeline uses the exact optical model used to generate the data, and a differentiable version of the exact $u,v$ transform implementation used in \textsc{xara}; we do not believe this to be the main cause of the small discrepancy between autodiff and conventional parameters, but it does illustrate that model mis-specification may be a contributing factor.

@@

We now consider a very simple Lyot coronagraph \citep{lyot30}, using \textsc{morphine}. We choose a 1\,$\mu$m wavelength, and 1\,m diameter pupil sampled on a $96\times96$ grid. The central $4 \lambda/D$ region of the resulting PSF is blocked out with a focal plane stop, and the light is then propagated to a second pupil plane using an FFT rather than a matrix FT, following standard \textsc{poppy} practice; this imposes tougher memory constraints, and in future may be replaced with an MFT. A Lyot stop (an iris mask undersized 10\% relative to the input pupil) is then imposed, and the light is propagated to a final focal plane sampled on a 50\,mas pixel scale grid with a 4~arcsec field of view. Aside from the coarse gridding, this is intentionally identical to one of the standard example pieces for \textsc{poppy}\footnote{\href{https://github.com/mperrin/poppy/blob/master/notebooks/MatrixFTCoronagraph_demo.ipynb}{github.com/mperrin/poppy/blob/master/notebooks/MatrixFTCoronagraph\_demo.ipynb}}, for the sake of consistency. The PSF produced has a dark coronagraphic hole in the middle, and beyond an inner working angle around this hole has a nearly normal Airy pattern of diffraction rings.

Again using \textsc{jax}, we calculate the Jacobian of this PSF with respect to the input wavefront, evaluated at uniform-zero phase, which again takes a few minutes on a laptop. The previous approach has been to directly propagate through the end to end simulation for a grid of small perturbations in the input plane \citep{falco}, whereas this is now taken care of by autodiff. The results are displayed in Figure~\ref{speckle_jacobian} in a similar format to Figure~\ref{kernel_jacobian}. Pixels far out in the PSF correspond to sinusoidal modulation of the input wavefront exactly as expected, and this is symmetric about the origin. On the other hand, pixels near the inner working angle where the effects of the coronagraph are evident correspond to interestingly-shaped modes in the pupil, and contain contributions from the part of the primary pupil that is obscured by the Lyot stop when the pupil is re-imaged.

Calculating the SVD of this Jacobian (again restricted to the support of the primary pupil) we se that unlike in the case of the simple diffraction, there is not a sharp cliff but instead a rolling-off which nevertheless reaches very small singular values. The singular value curve is displayed in Figure~\ref{fig:svd_coronagraph}.

@@

For the coronagraph under consideration, the linear regime may be of limited validity. We take one of the sinusoidal ripples which are the columns of the Jacobian in Figure~\ref{speckle_jacobian}, propagate this through a full optical simulation, and take the difference with respect to an unaberrated PSF. In Figure~\ref{corona_nonlinearity}, we see the effects of a sinusoidal phase ripple in the exactly linear regime, with an 0.1~nm amplitude, and a 5.0~nm amplitude. In the linear regime and where the phase ripple is small, we see that the speckle is positive on one side of the PSF and negative on the other. On the other hand, where the phase ripple is even a few nanometres, we see that the speckles are both positive: an indication of the quadratic term in the Taylor series becoming dominant. This quadratic term involves a rank-3 tensor, of the second partial derivatives of each pixel with respect to the wavefront - and there is no equivalent of a kernel operator for tensors of this dimension. The kernel phase trick will therefore not work straightforwardly in this case, though we suggest that future work may search for a higher-dimensional kernel manifold.

We have shown that self-calibrating observables exist for coronagraphs. But this nonlinearity argument indicates they may be of limited practical applicability to coronagraphic observations except in the extremely high-wavefront-quality regime. Sub-nanometre precision may nevertheless be achieved in future large space telescopes with high-order deformable mirrors.

@@

A phase transfer matrix constructed analytically with a discrete pupil model does not know a priori about the pixel sampling, binning, or windowing. This means that there is convolution in the $u,v$ planeand therefore correlation between adjacent baselines, which means that the normally-orthonormal kernel phases (calculated analytically for an infinite field of view and fine sampling) are no longer linearly independent. For example \citet{martinache20} notes that in the analysis of \citet{palomar}, windowing to avoid the ghost introduced by a neutral density filter cuts the number of pixels to less than the number of kernel phases, so that information is being lost. \citet{martinache20} also demonstrates how using a detailed model of pupil (mis-)alignment can improve the extraction of both kernel phase and visibility information.

@@

By calculating kernel phases from differentiating an end-to-end optical simulation, even for trivial apertures, you can include the exact pixel grid, including binning, windowing, and dead pixels. As the matrix DFT is already used to calculate the FT, there is no sampling noise to separately account for, and all kernel phases are automatically orthonormal.

% forwards model using these matrices a la Pueyo
To avoid self-subtraction, we expect that the best approach to data analysis with kernel phases using these methods will be forwards-modelling the data through the optical system and kernel phase filter along the lines described by \citet{pueyo16} or \citet{martinache20}. These can then be augmented by diversity-based methods along the lines described by \citet{ireland13}, to achieve even higher levels of calibration. We have not attempted to use these methods on real data in this paper.

@@

This idea originated in a conversation with Laurent Pueyo and Nour Skaf. I would like to thank them together with Yinzi Xin, Anand Sivaramakrishnan, Marshall Perrin, Will Farr, David Hogg, and Peter Tuthill for their very helpful comments.

>>>>

\author[0000-0003-3818-408X]{Laurent Pueyo}
\affiliation{Space Telescope Science Institute, 3700 San Martin Drive, Baltimore, MD, 21218, USA}

\author[0000-0002-6171-9081]{Yinzi Xin}
\affiliation{Department of Aeronautics and Astronautics, Massachusetts Institute of Technology, Cambridge, MA 02139, USA}

\author[0000-0001-7026-6291]{Peter~G.~Tuthill}
\affiliation{Sydney Institute for Astronomy (SIfA), School of Physics, The University of Sydney, NSW 2006, Australia}

%% Mark off the abstract in the ``abstract'' environment. 
\begin{abstract}
% optical imaging with aberrations
The accumulation of aberrations along an  optical path produces distortions and speckles in the resulting images, limiting the performance of cameras at high angular resolution.

% self cal - closure phase and kernel phase
Self-calibrating observables such as the `closure phase' or `bispectrum' have been widely used in optical and radio astronomy to mitigate these issues: a linear combination of measurements with phase noise can be made such that error terms cancel, leaving a subspace of high signal-to-noise observables. Kernel phases are a generalization of closure phases, in which the effect of aberration on an arbitrary pupil is approximated as a linear phase transfer matrix valid in the limit of small phase errors.
%coronagraphs are hard
Finding this operator is analytically tractable for simple optical paths (such as propagation from a  single pupil to focus) but is much more difficult for other optical systems of interest, such as coronagraphs.
% autodiff can make this practical for arbitrary optical systems
Automatic differentiation or `backpropagation' software now makes calculating derivatives with respect to aberrations in arbitrary planes straightforward for any optical system.
% we demonstrate a simple example of the kernel of a monochromatic coronagraph
We reproduce existing kernel phase theory within this framework and demonstrate an extension to the case of a Lyot coronagraph, which is found to have self-calibrating combinations of speckles which are resistant to phase noise, but only in the very high wavefront quality regime. We also use this framework to reanalyze Palomar adaptive optics observations of the binary $\alpha$~Ophiuchi, finding broad consistency and slight systematics between the new pipeline and the existing standard.
%phase design
We present a Python package \textsc{morphine} that incorporates these ideas, with an interface similar to the popular package \textsc{poppy}, for optical simulation with automatic differentiation. These methods may be useful for designing improved astronomical optical systems by gradient descent.
\href{https://github.com/benjaminpope/morphine}{\color{linkcolor}\faGithub} % shamelessly borrowed from Luger!

@@

\section{Introduction} 
\label{sec:intro}
% optical imaging with aberrations
Many questions in astronomy and other sciences can only be answered with diffraction-limited high resolution imaging. The highest resolutions are typically achieved with the method of interferometry, in which waves detected at multiple receivers are combined physically or in post-processing to obtain the Fourier transform of the source intensity distribution \citep{vc34,zernike38}. Even in the case of a single telescope or camera, it is often nevertheless helpful to think of them as an interferometer composed of many sub-apertures which combine their signals directly onto a focal plane \citep[a `Fizeau interferometer':][]{fizeau1868}. In all of these cases, unknown path delays toward each receiver or subaperture are a dominant source of noise, causing `tip-tilts' of the final image position, distortions of the point spread function (PSF) from low-spatial-frequency aberrations, and clouds of `speckles' from high-spatial-frequency aberrations.

% self cal with closure phase
In the recovery of object phase information, among the most longstanding techniques is that of self-calibration with `closure phases' \citep[introduced in the context of radio astronomy by][]{jennison58}, in which phases are summed around three interferometric baselines which form a closing triangle. The phase error terms local to each subaperture cancel, so that three low-signal-to-noise (SNR) baseline phases deliver one high-SNR observable. In a non-redundant array (one in which no baseline vector is repeated between different pairs of subapertures) of sufficient size, a large number of closure phases can be obtained which anchor image reconstruction with great precision. `Closure amplitudes' which are resistant to fluctuations in input amplitude or gain can also be obtained by a similar construction using four telescopes \citep{twiss60,blackburn20}. The closure phase is the argument of a quantity called the triple product or `bispectrum', and it has recently been shown from the perspective of invariant theory in algebraic geometry that for a wide class of problems limited by phase noise, knowledge of the mean, power spectrum and bispectrum are necessary and sufficient for an optimal signal reconstruction \citep{bandeira17}.

On the other hand, for the case of direct imaging of high contrast companions with coronagraphs, analytic self calibrations such as those above are not yet known. The standard approaches to data analysis in coronagraphy rely on exploring a diversity of PSFs experimentally, constructing a basis covering some of the diversity in speckle patterns, and then subtracting out a linear combination of vectors in this basis from measured data \citep[e.g.][]{lafreniere07,soummer12,pueyo16}. Advantages can also be gained from angular differential imaging \citep[ADI;][]{marois06} and spectral deconvolution with wavelength \citep{sparks02}. In this paper we will extend our understanding of analytic self calibration to better include coronagraphs, so that we can add an additional layer of precision to exoplanet imaging calibration.

@@

% coronagraphs, speckles etc
It is desirable to find a priori self-calibration schemes for high-angular-resolution, high-contrast imaging systems more generally, including not merely nulling interferometers but optically complex coronagraphs intended to suppress light in a region around a target star. One particularly interesting extension of the kernel phase idea is the `kernel nuller' concept \citep{martinache18}, in which the aberration transfer matrix and kernel operator idea is applied to the output of a nulling interferometer.

@@

The \textsc{DeepOptics} project \citep{sitzmann2018} has used autodiff to optimize `computational cameras'. Building their model in \textsc{TensorFlow}, they couple a physical optics simulation, detector simulation, and a deconvolution post-processing stage for a total end-to-end imaging simulation. Where normally you might optimize some heuristic of the PSF (such as the full width at half maximum), this makes it possible to jointly optimize hardware and software with respect to figures of merit of the overall system such as final resolution or depth of field. Some designs arrived at in this way are exotic: for example, super-resolution is achieved by finding a lens with three off-centre Fresnel-lens components focusing to three separate spots, which the deconvolution stage shifts and stacks. Because it is built in \textsc{TensorFlow}, the diffraction simulation can be incorporated as a `physical layer' in neural network applications in microscopy, for example for optimizing hardware and software for image classification \citep{muthumbi19}, or with reinforcement learning for adaptively learning sample illumination \citet{chaware19}.

@@

Rather than using these autodiff gradients for optimization, in this paper we use them to understand optical systems and the information they propagate.
In the following we will show how autodiff can reproduce the existing state of the art in kernel phase, and demonstrate a way forward using autodiff to extend the kernel phase idea to coronagraphic instruments.

@@

For our forwards model, we adapt the popular physical optics library \textsc{poppy} \citep{poppy}, with the goal of compatibility with existing simulations built on \textsc{poppy} such as \textsc{WebbPSF} \citep{webbpsf}. We use the matrix Fourier transform mode to avoid using the FFT \citep{soummer07}, which is more efficient for autodiff applications. We have adapted the low-level features of \textsc{poppy} to use the Google autodiff library \textsc{jax} \citep{jax} in place of \textsc{NumPy}, and to distinguish it from the original version we call this new \textsc{poppy} derivatives library \textsc{morphine}. %The \textsc{NumPy}-like API of \textsc{jax} is chosen so that astronomers can painlessly integrate \textsc{morphine} with their existing code, but we note that a \textsc{DeepOptics}-like approach using \textsc{TensorFlow} may be preferable in the longer term for integration with neural networks.

@@

First we want to reproduce the basic kernel phase calculations in this new model. We propagate monochromatic 2.0\,$\mu$m light through a 2.0\,m diameter circular pupil onto a detector with a 20\,mas/pixel scale and 4\,arcsec field of view. The pupil and $u,v$ are each calculated on a 64x64 grid. We evaluate the Jacobian of the $u,v$ phases with respect to the input pupil phases using forwards-mode autodiff and then evaluate this Jacobian at zero phase to get our $\mathbf{A}$ matrix\footnote{\href{https://github.com/benjaminpope/morphine/blob/master/notebooks/morphine_u,v.ipynb}{github.com/benjaminpope/morphine/blob/master/notebooks/morphine\_u,v.ipynb}}. On a laptop computer this takes a few minutes processing time. Memory usage is the main overhead: in practice on a laptop with 16GB RAM, we are limited to grids no larger than 128x128. Demand on computational resources presents a major limitation to widespread application; in future this may be overcome by more efficient refactoring of the code, or simply by doing calculations on clusters with very large RAM allocations.

Although it is inefficient to calculate this full Jacobian with respect to a densely-pixelized basis for the wavefront in the pupil plane, we do so here for completeness. In practical cases, aberrations are typically dominated by low-order modes and we could consider the Jacobian only with respect to (for example) some much smaller number of Zernike modes. Reducing the input dimensionality like this would have significant advantages in memory usage for forwards-mode autodiff.

@@

We then extract kernel phases from both $\alpha$~Oph and a point source calibrator $\epsilon$~Her using a version of \textsc{xara} modified to use the \textsc{morphine} matrix FT. Following \citet{martinache20}, we choose as our observables the median kernel phases across all 100 exposures for each, and the standard error of the mean as our base uncertainty on each. To calibrate the kernel phases, we simply subtract those of the calibrator from $\alpha$~Oph and add their uncertainties in quadrature.

To see the effect of the known optical `ghost' (introduced by an unwanted reflection from the neutral density filter used to observe bright stars), we calculate kernel phase `colinearity' maps as in \citet{martinache20}, shown in Figure~\ref{colinearity}. In these maps we can see that the filter ghost shows up strongly as a false binary in the uncalibrated kernel phase maps, but calibrated kernel phases easily remove this and reveal the tight $\alpha$~Oph binary.

@@

We then use Markov Chain Monte Carlo \citep{metropolis53} to sample from the posterior assuming Gaussian likelihoods, using the \texttt{emcee} affine-invariant ensemble sampler \citep{emcee}. We obtain binary parameters of separation $\rho = 115.3 \pm 1.2$~mas, position angle $\theta = 85.6 \pm 0.27$~degrees, and contrast $c = 26.3 \pm 0.6$: these are close to the \citet{martinache20} figures, but differ by much more than their (small) uncertainties. For comparison, the contrast ratio of the system is expected from resolved imaging to be $28.0 \pm 8.3$ in $K$-band \citep{hinkley11}. Posteriors for both inferences are shown in Figure~\ref{comparison_posterior}.

@@

In order to investigate this discrepancy, we generated a grid of forwards models. Using the same code used to produce the kernel phase autodiff model, we generated input wavefronts with 20 Zernike modes of aberrations, each with random amplitudes normally-distributed with 15\,nm standard deviation, for a total peak-to-peak amplitude of 206\,nm. These were then shifted and added to simulate a range of binaries at different configurations and contrasts, and both the \citet{martinache20} pipeline and the autodiff pipeline were used to retrieve their parameters. The results are shown in Figure~\ref{method_comparison_ensemble}. The pipelines achieve broadly comparable performance on these simulations, with the \citet{martinache20} method being slightly better at closer separations and higher contrasts and autodiff performing marginally better at larger separations.

@@

\begin{figure}
\plotone{comparison_posterior_sim.pdf}
\caption{Comparison of \citet[][blue linetype]{martinache20} and autodiff (black) kernel phase pipelines on simulated data with true parameters $\rho = 125$~mas, $\theta = 86^\circ$, $c=25$. We generate a cube of 1000 mock datasets using a 64-pix pupil model and Zernike aberrations as described in Section~\ref{sec:palomar}, and the observables used are the median and standard error of the mean of each kernel phase across this ensemble. Both datasets retrieve the true parameters very precisely, though with small systematic offsets, which are slightly larger for the autodiff method than for classic kernel phases. \label{comparison_posterior_sim}}
\end{figure}

\begin{figure}
\plotone{comparison_posterior_sim_cube.pdf}
\caption{Comparison of \citet[][blue linetype]{martinache20} and autodiff (black) kernel phase pipelines on simulated data with true parameters $\rho = 125$~mas, $\theta = 86^\circ$, $c=25$. We generate a cube of 1000 mock datasets using a 64-pix pupil model and Zernike aberrations as described in Section~\ref{sec:palomar}. The corner plot displays the parameters retrieved using a simple least squares fit applied to each individual frame. The methods have very similar distributions centred tightly on the true value. \label{comparison_posterior_sim_cube}}
\end{figure}

In numerical experiments, subpixel mismatch in Fourier transform coordinates or even Fourier centering convention could cause mismatches between simulation parameters and retrievals at around the scale of the discrepancy between the two methods.  These results are in line with the \citet{martinache_habilitation} conclusion that inexact Fourier sampling contributes significant systematic noise to kernel phases. The final version used for the autodiff pipeline uses the exact optical model used to generate the data, and a differentiable version of the exact $u,v$ transform implementation used in \textsc{xara}; we do not believe this to be the main cause of the small discrepancy between autodiff and conventional parameters, but it does illustrate that model mis-specification may be a contributing factor.

@@

We now consider a very simple Lyot coronagraph \citep{lyot30}, exploiting the ability of \textsc{morphine} to analyze systems that would otherwise be too computationally demanding. We choose a 1\,$\mu$m wavelength, and 1\,m diameter pupil sampled on a $96\times96$ grid. The central $4 \lambda/D$ region of the resulting PSF is blocked out with an on-axis focal plane stop (occulter), and the light is then propagated to a second pupil plane using an FFT rather than a matrix FT, following standard \textsc{poppy} practice. This imposes tougher memory constraints, and in future may be replaced with an MFT. A Lyot stop (an iris mask undersized 10\% relative to the input pupil) is then imposed, and the light propagated to a final focal plane sampled on a 50\,mas pixel scale grid with a 4~arcsec field of view. Aside from the coarse gridding, this is intentionally identical to one of the standard test and verification examples supplied with \textsc{poppy}\footnote{\href{https://github.com/mperrin/poppy/blob/master/notebooks/MatrixFTCoronagraph_demo.ipynb}{github.com/mperrin/poppy/blob/master/notebooks/MatrixFTCoronagraph\_demo.ipynb}}. The PSF produced has a dark coronagraphic hole in the middle, and beyond an inner working angle around this hole has a pattern of diffraction rings.

Again using \textsc{jax}, we calculate the Jacobian of this PSF with respect to the input wavefront, evaluated at uniform-zero phase, which again takes a few minutes on a laptop. The previous approach has been to directly propagate through the end to end simulation for a grid of small perturbations in the input plane \citep{falco}, whereas this is now taken care of by autodiff. The results are displayed in Figure~\ref{speckle_jacobian} in a similar format to Figure~\ref{kernel_jacobian}. Pixels far out in the PSF correspond to sinusoidal modulation of the input wavefront as expected, and this is symmetric about the origin. On the other hand, pixels near the inner working angle where the effects of the coronagraph are evident correspond to more complex modes in the pupil, also containing contributions from the ring obscured by the Lyot stop when the pupil is re-imaged.

Calculating the SVD of this Jacobian (again restricted to the support of the primary pupil) we see a decline to very small singular values, albeit not as steep as earlier for the simple imaging case. The ordered singular value curve is displayed in Figure~\ref{fig:svd_coronagraph}.

@@

For the coronagraph under consideration, the regime over which the linear approximation applies may strictly limited. To illustrate this, we take one of the sinusoidal ripples which are the columns of the Jacobian in Figure~\ref{speckle_jacobian}, propagate this through a full optical simulation, and take the difference with respect to an unaberrated PSF. In Figure~\ref{corona_nonlinearity}, we see the effects of a sinusoidal phase ripple in the nearly-exact linear regime with a 0.1~nm amplitude, and a with a small but non-negligible 5.0~nm amplitude. In the linear regime and where the phase ripple is very small, we see that the speckle is positive on one side of the PSF and negative on the other corresponding closely with expected behavior from the Jacobian. On the other hand increaasing the amplitude of the phase ripple is even to  5 nanometres is enough to change the outcome completely. We see from the figure that the speckles are now both positive: a radical switch in behavoir indicating the quadratic term in the Taylor series has become dominant. This quadratic term involves a rank-3 tensor, of the second partial derivatives of each pixel with respect to the wavefront -- and there is no equivalent of a kernel operator for tensors of this dimension. The use of kernel phases as robust observables will therefore fail in this case, though we suggest that future work may search for a higher-dimensional kernel manifold.

We have therefore shown that under the same linear assumptions previously applied to yield kernel phases, analogous self-calibrating observables also exist for propagation through more complex optical systems such as coronagraphs. However for the example configuration explored, the rapid onset of non-linear response indicates they may be of limited practical applicability to coronagraphic observations except in the extremely high-wavefront-quality regime. Sub-nanometre wavefront precision may nevertheless be achieved in future large space telescopes with high-order deformable mirrors.

@@

A phase transfer matrix constructed analytically with a discrete pupil model contains no a priori information about the pixel sampling, binning, or windowing. This means that the implicit convolutions in the $u,v$ plane, and therefore the associated correlation between adjacent baselines, is not taken into account.  Therefore the normally-orthonormal kernel phases (calculated analytically for an infinite field of view and fine sampling) are no longer fully linearly independent. For example \citet{martinache20} notes that in the analysis of \citet{palomar}, windowing to avoid the ghost introduced by a neutral density filter cuts the number of pixels to less than the number of kernel phases, so that information is being lost. \citet{martinache20} also demonstrates the utility of a detailed model of pupil (mis-)alignment to improve the extraction of both kernel phase and visibility information.

@@

By calculating kernel phases from differentiating an end-to-end optical simulation, even for trivial apertures, the exact pixel grid, including binning, windowing, and dead pixels can be included. As the matrix DFT is already used to calculate the FT, there is no sampling noise to separately incorporate, and all kernel phases are automatically orthonormal.

% forwards model using these matrices a la Pueyo
To avoid self-subtraction, we expect that the best approach to data analysis with kernel phases using these methods will be forwards-modelling the data through the optical system and kernel phase filter along the lines described by \citet{pueyo16} or \citet{martinache20}. These can then be augmented by diversity-based methods along the lines described by \citet{ireland13} to achieve even higher levels of calibration. We have not attempted to use these methods on real data in this paper.

@@

We would like to thank Anand Sivaramakrishnan, Marshall Perrin, Will Farr, and David Hogg for their very helpful comments.
